---
title: "Assignment 3: Classification Trees and Bagging"
author: "Millie Spencer"
output: html_document
---

**Due:** Thursday 19 Feb 3:30 PM

## Learning goals

* Understand the steps of the bagging model algorithm
* Practice modifying a function to examine its internal workings
* Understand the differences between regression and classification algorithms
* Practice tuning machine learning algorithms using the CV inference algorithm
* Practice comparing model accuracy using the CV inference algorithm
* Understand how the bagging algorithm averages over probabilities in classification models
* Practice visualizing model outputs

## Classification trees and bagging

The goal of this assignment is to compare the predictive performance (using k-fold CV) of a decision-tree model and a bagged decision tree model to the KNN model from the previous assignment for the presence/absence of the Willow Tit. We will again use the single predictor, elevation.

**Q1\.** Train a classification tree. For R, use the `tree()` function from the `tree` package. The option `type="class" ` in `predict()` will return the predicted class (see `?predict.tree`). Alternatively, for Python, use the `tree.DecisionTreeClassifier()` function from the [scikit-learn](https://scikit-learn.org) package (see, e.g. in `ants_bag.py`). Tune the tree for the minimum number of samples to split a node (R: `mincut`; Python: `min_samples_split`) using k-fold cross validation. Feel free to share your tuning plot on Piazza and see what others are getting.
```{r Q1 Tree}
# packages
library(ggplot2)
library(dplyr)
library(tree)

# Load willowtit data (setting presence =1 and absence =0)
willowtit <- read.csv("data/wtmatrix.csv") |> 
    mutate(status = ifelse(y.1 == 1, "present", "absent")) |>
    mutate(status = factor(status)) |>   
    select(status, elev)

# define the random_partitions function 
random_partitions <- function(n, k) {
    # Creates k random partitions of numbers 1 to n
    partitions <- rep(1:k, length.out = n)
    sample(partitions)
}  # <-- closing brace was missing before!

# Tune mincut using k-fold CV
# mincut controls minimum number of observations allowed in a terminal node
# (similar to how we tuned k_knn last time)
cv_tree_willowtit <- function(willowtit, k_cv, mincut) {
    willowtit$partition <- random_partitions(nrow(willowtit), k_cv)
    e <- rep(NA, k_cv) # e is the vector inside the function that stores the error rate for each fold. For example with k_cv = 10, e has 10 slots — one error rate per fold — and at the end the function averages them all into one number and returns it.
    for (i in 1:k_cv) {
        test_data <- subset(willowtit, partition == i)
        train_data <- subset(willowtit, partition != i)
        # Train tree with this mincut value
        trained_tree <- tree(status ~ elev, data = train_data,
                            control = tree.control(nobs = nrow(train_data), 
                                                    mincut = mincut))
        # Predict classes (present or absent) on test data
        pred_status <- predict(trained_tree, newdata = test_data, type = "class")
        # Classification error rate
        e[i] <- mean(test_data$status != pred_status) # calculates the error rate for this fold, compares the real labels to the predicted labels and returns TRUE/FALSE for each row — TRUE means the prediction was wrong (aka prediction is not equal to (!) reality)
    }
    return(mean(e)) 
}

```

```{r Q1-pt2 tuning, cache=TRUE}
# Define the mincut values we want to test - these control how many observations a node needs before the tree will try to split it further.
# Small mincut = more splits = more complex tree (but risk of overfitting)
# Large mincut = fewer splits = simpler tree
mincut_values <- c(1, 5, 10, 15, 20, 25, 30, 35, 40, 50, 75, 100)

# Create an empty results table with one row per mincut value
cv_results <- data.frame(mincut = mincut_values, cv_error = NA)

# Repeat the CV process 200 times. Each repeat uses different random partitions, so averaging across all reps gives a much more stable estimate of the true CV error
reps <- 200

# Matrix to store results: rows = mincut values, columns = replicate runs
cv_error_matrix <- matrix(NA, nrow = length(mincut_values), ncol = reps)

set.seed(123) # makes results reproducible

# Outer loop: repeat the whole CV process 200 times
for (j in 1:reps) {
    # Inner loop: test each mincut value within this replicate
    for (i in seq_along(mincut_values)) {
        cv_error_matrix[i, j] <- cv_tree_willowtit(willowtit, 
                                                    k_cv = 10, 
                                                    mincut = mincut_values[i])
    }
}

# Average the 200 CV error estimates for each mincut value. This smooths out the noise from random partitioning
cv_results$mean_cv_error <- rowMeans(cv_error_matrix)

# Find the mincut with the lowest average CV error
best_mincut <- cv_results$mincut[which.min(cv_results$mean_cv_error)]
cat("Best mincut:", best_mincut, "\n")
print(cv_results)

# Plot mean CV error vs mincut. Looking for a clear minimum - error should go down then back up
ggplot(cv_results, aes(x = mincut, y = mean_cv_error)) +
    geom_line() +
    geom_point() +
    geom_vline(xintercept = best_mincut, linetype = "dashed", color = "red") +
    labs(title = "Mean CV Error vs mincut (averaged over 200 runs)",
         x = "mincut", 
         y = "Mean CV Error Rate") +
    theme_minimal()

# Train the final classification tree using the best mincut value
final_tree <- tree(status ~ elev, data = willowtit,
                   control = tree.control(nobs = nrow(willowtit), 
                                          mincut = best_mincut))
summary(final_tree)
```

**Q2\.** Plot predictions (as a map) from the classification tree. The code for this will be substantially similar to what you did in the previous assignment.
```{r Q2- plot predictions from classification tree}
# Generate predictions for all Swiss elevations using our tuned tree
# Need to rename Elev_m to elev to match the column name the tree was trained on
swiss_predictions <- predict(final_tree, 
                             newdata = rename(swissdem, elev = Elev_m), 
                             type = "class")

# Add predictions to swissdem
swissdem$pred_status <- swiss_predictions

# Map predictions
ggplot() +
    geom_raster(data = swissdem,
                aes(x = x, y = y, fill = Elev_m)) +
    scale_fill_gradient(low = "white", high = "black", name = "Elevation (m)") +
    geom_tile(data = filter(swissdem, pred_status == "present"),
              aes(x = x, y = y, color = "Predicted Presence"),
              fill = "magenta", alpha = 0.6, linewidth = 0) +
    scale_color_manual(name = "",
                       values = c("Predicted Presence" = "magenta")) +
    coord_quickmap() +
    labs(title = "Classification Tree: Predicted Willow Tit Distribution") +
    theme_void()
```

**Q3\.** Modify the `bagrt()` function to make a `bagct()` function for bagged classification trees. This is very similar to the step we made from regression to classification for the KNN model (look at `classification_knn.R/py` for inspiration). Whereas for bagged regression trees we averaged the prediction for a numerical response variable (e.g. richness in the ants example) across trees, for bagged classification trees we need to average the probabilities across trees before making the final present/absent prediction (equivalently we can vote across trees).  For R, the option `type="vector" ` in `predict()` will return the probabilities (see `?predict.tree`). For Python, the function `tree.predict_proba()` will return the probabilities (see [sklearn API](https://scikit-learn.org/stable/modules/generated/sklearn.tree.DecisionTreeClassifier.html#sklearn.tree.DecisionTreeClassifier.predict)). 
```{r Q3- Bagged classification tree function}
# Modified from bagrt() to handle classification (present vs not present) instead of regression
# Instead of averaging numeric predictions, we average probabilities across trees then threshold at 0.5 to get present/absent

# formula:    model formula (formula)
# data:       y and x data (data.frame)
# xnew_data:  x data to predict from (data.frame)
# boot_reps:  number of bootstrap replications (here trying 500)
# return:     predicted classes (vector, character)

bagct <- function(formula, data, xnew_data, boot_reps = 500) {
    n <- nrow(data) # 237 willow tit observations
    nx <- nrow(xnew_data) # however many points we want predictions for
    # Matrix to store probability of presence from each tree
    boot_probs <- matrix(NA, nrow = nx, ncol = boot_reps) # creates empty matrix, one row per prediction point and one column per bootstrap tree (eg matrix w 500 columns)
    for (i in 1:boot_reps) {
        # Resample the data with replacement
        boot_indices <- sample(1:n, n, replace = TRUE) # randomly pick 237 rows WITH replacement - some rows appear twice, some not at all
        boot_data <- data[boot_indices, ]
        # Train one classification tree on the bootstrap sample
        boot_train <- tree(formula, data = boot_data)
        # Record probability of presence 
        boot_probs[, i] <- predict(boot_train, newdata = xnew_data, 
                                   type = "vector")[, "present"]
    }
    # average across all 500 trees for each prediction point
    # e.g. if 400 trees said 0.8 probability and 100 said 0.2, average = 0.68
    avg_probs <- rowMeans(boot_probs)
    # finally convert that average probability to a class label
    final_pred <- ifelse(avg_probs > 0.5, "present", "absent")
    return(final_pred)
}
```

```{r Quick test - run bagct with a small number of boot_reps to check it works}
test_pred <- bagct(status ~ elev, 
                   data = willowtit, 
                   xnew_data = rename(swissdem, elev = Elev_m), 
                   boot_reps = 10)

# Check the output looks sensible - should be a vector of "present"/"absent"
head(test_pred)
table(test_pred)
# yay! like before predicting presence in mid-elevation band 
```

**Q4\.** Tune `bagct()` for the number of bootstrap samples on the Willow Tit data. Yep, you know the gig by now: use k-fold CV. Feel free to share your tuning plot on Piazza and see what others are getting.

```{r Q4-cv-function}
# CV function for bagged classification tree
# Same structure as cv_tree_willowtit but calls bagct() instead
# willowtit:  dataset
# k_cv:       number of folds
# boot_reps:  number of bootstrap samples - this is what we're tuning
# return:     CV error rate
cv_bagct <- function(willowtit, k_cv, boot_reps) {
    willowtit$partition <- random_partitions(nrow(willowtit), k_cv)
    e <- rep(NA, k_cv)
    for (i in 1:k_cv) {
        test_data <- subset(willowtit, partition == i)
        train_data <- subset(willowtit, partition != i)
        # Call bagct on training data, predict for test data
        pred_status <- bagct(status ~ elev,
                             data = train_data,
                             xnew_data = test_data,
                             boot_reps = boot_reps)
        # Classification error rate
        e[i] <- mean(test_data$status != pred_status)
    }
    return(mean(e))
}
```
```{r Q4-tuning, cache=TRUE}
# Tune boot_reps using k-fold CV
# Brett suggested optimum is probably less than 50
# Testing a range focused around that area
boot_reps_values <- c(5, 10, 20, 30, 40, 50, 75, 100)

# Empty results table
boot_results <- data.frame(boot_reps = boot_reps_values, mean_cv_error = NA)

# Matrix to store results across replicate runs
# Fewer reps than before because each run is slower (bagct loop inside CV loop!)
reps <- 50
cv_error_matrix_bag <- matrix(NA, nrow = length(boot_reps_values), ncol = reps)

set.seed(123)
for (j in 1:reps) {
    for (i in seq_along(boot_reps_values)) {
        cv_error_matrix_bag[i, j] <- cv_bagct(willowtit,
                                               k_cv = 10,
                                               boot_reps = boot_reps_values[i])
    }
}

# Average across replicate runs
boot_results$mean_cv_error <- rowMeans(cv_error_matrix_bag)
best_boot_reps <- boot_results$boot_reps[which.min(boot_results$mean_cv_error)]
cat("Best boot_reps:", best_boot_reps, "\n")
print(boot_results)
# lowest is 75 but really tiny difference between boot reps

# Plot - looking for where error stops decreasing meaningfully
ggplot(boot_results, aes(x = boot_reps, y = mean_cv_error)) +
    geom_line() +
    geom_point() +
    geom_vline(xintercept = best_boot_reps, linetype = "dashed", color = "red") +
    labs(title = "Mean CV Error vs boot_reps (averaged over 50 runs)",
         x = "Number of bootstrap samples",
         y = "Mean CV Error Rate") +
    theme_minimal()
```

**Q5\.** Plot predictions (as a map) from the tuned, bagged classification tree algorithm.

```{r Q5 - Train bagged classification tree with best boot_reps}
swiss_bag_predictions <- bagct(status ~ elev,
                                data = willowtit,
                                xnew_data = rename(swissdem, elev = Elev_m),
                                boot_reps = 50)

# Add predictions to swissdem
swissdem$pred_status_bag <- swiss_bag_predictions

# Map predictions
ggplot() +
    geom_raster(data = swissdem,
                aes(x = x, y = y, fill = Elev_m)) +
    scale_fill_gradient(low = "white", high = "black", name = "Elevation (m)") +
    geom_tile(data = filter(swissdem, pred_status_bag == "present"),
              aes(x = x, y = y, color = "Predicted Presence"),
              fill = "magenta", alpha = 0.6, linewidth = 0) +
    scale_color_manual(name = "",
                       values = c("Predicted Presence" = "magenta")) +
    coord_quickmap() +
    labs(title = "Bagged Classification Tree: Predicted Willow Tit Distribution") +
    theme_void()
```

**Q6\.** Compare predictive accuracy of KNN, simple tree, and bagged tree models for the Willow Tit dataset by collating the CV results. Which model is best?
```{r Q6 - Compare models via CV results}
# Collate CV errors from all three models for comparison
# from HW 2, LOOCV yieleded the lowest error value (22.4%) at a K of between 40-44

model_comparison <- data.frame(
    model = c("KNN (k=41)", "Single Tree (mincut=35)", "Bagged Tree (boot_reps=50)"),
    cv_error = c(0.224,                              # LOOCV from assignment 2
                 min(cv_results$mean_cv_error),      # from Q1 tuning = 0.235
                 min(boot_results$mean_cv_error))    # from Q4 tuning = 0.242
)

print(model_comparison)
cat("\nBest model:", model_comparison$model[which.min(model_comparison$cv_error)], "\n")
```

Interesting, KNN wins with the lowest error (22.4%), followed by the single tree (23.5%), then bagged tree (24.2%)! Surprised that KNN wins. 
because it can capture the smooth bell-shaped relationship between elevation and presence more naturally than a tree-based approach. The single tree and bagged tree 
are very close to each other, consistent with Brett's observation of only ~1% improvement from bagging on this dataset.

**Q7\.** Modify the code for your **best-performing model** to output the probabilities instead of presence/absence class. Plot the probabilities as a map.
```{r Q7 - best model (KNN) probabilities of willow tit presence}
# Generate probability predictions for Switzerland using best KNN model (k=41)
# Based on script from last assignment - used knn_classify2_probs on a small elevation grid just to visualize the probability curve. Here we're applying it to the full Swiss map.

# Probability version of KNN classify - returns probability of presence
# instead of class labels. Modified from knn_classify2 by removing the
# thresholding step and always returning probability of "present"
knn_classify2_probs <- function(x, y, x_new, k) {
    category <- unique(y)
    y_int <- ifelse(y == category[1], 1, 0)
    nx <- nrow(x)
    n <- nrow(x_new)
    c <- ncol(x_new)
    p_cat1 <- rep(NA, n)
    for ( i in 1:n ) {
        x_new_m <- matrix(x_new[i,], nx, c, byrow=TRUE)
        d <- sqrt(rowSums((x - x_new_m) ^ 2))
        y_sort <- y_int[order(d, sample(1:length(d)))]
        p_cat1[i] <- mean(y_sort[1:k])
    }
    if (category[1] == "present") {
        return(p_cat1)
    } else {
        return(1 - p_cat1)
    }
}

swiss_probs <- knn_classify2_probs(
    x = as.matrix(willowtit["elev"]),
    y = willowtit$status,
    x_new = as.matrix(swissdem["Elev_m"]),
    k = 41
)

# Add probabilities to swissdem
swissdem$prob_present <- swiss_probs

# Map probabilities as a continuous color scale instead of binary present/absent
ggplot(swissdem, aes(x = x, y = y, fill = prob_present)) +
    geom_raster() +
    scale_fill_gradient(low = "white", high = "magenta", 
                        name = "Probability\nof Presence") +
    coord_quickmap() +
    labs(title = "KNN (k=41): Probability of Willow Tit Presence") +
    theme_void() +
    theme(
        plot.title = element_text(hjust = 0.5, vjust = -2, face = "bold"),
        legend.position = c(0.95, 0.95),
        legend.justification = c(1, 1),
        legend.background = element_rect(fill = "white", color = "black")
    )
summary(swissdem$prob_present)
```